---
title: "Class 12"
author: "Erin Li"
format: html
---
Q5
```{r}
# read the file
genotypes <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (2).csv")

head(genotypes)
nrow(genotypes)

# Subset for MXL population
mxl_genotypes <- genotypes[grepl("MXL", genotypes$Population),]

# Calculate the frequency of the G|G genotype
gg_frequency <- table(mxl_genotypes$Genotype)

# Calculate the proportion of G|G genotype
gg_proportion <- gg_frequency["G|G"] / sum(gg_frequency)

# Print the proportion
print(gg_proportion)
```

```{r}
file_path = '/mnt/data/rs8067378_ENSG00000172057.6.csv'

# read the file
data <- read.table("rs8067378_ENSG00000172057.6.csv", header = TRUE, sep = "")

# Display the first few rows of the dataframe to understand its structure
head(data)
```
```{r}
library(dplyr)
summary_data <- data %>%
  group_by(geno) %>%
  summarise(SampleSize = n(),
            MedianExpression = median(exp, na.rm = TRUE))
```
Q13
```{r}
# Print the summary data
print(summary_data)
```
Q14
```{r}
boxplot(exp ~ geno, data = data, main = "Expression Levels by Genotype",
        xlab = "Genotype", ylab = "Expression Level")
```
```{r}
library(ggplot2)

ggplot(data, aes(x=geno, y=exp, fill=geno)) + 
  geom_boxplot() +
  geom_jitter(shape=16, position=position_jitter(0.2), alpha=0.5) + # Adds the points on the plot
  scale_fill_manual(values=c("#FF0000","#009E73", "#56B4E9")) + # Change colors as needed
  labs(x="Genotype", y="Expression") +
  theme_minimal()
```
The median expression level for the genotype A/A is higher than for G/G.
The interquartile range (IQR), which represents the middle 50% of the data, is narrower for G/G than for A/A, suggesting that expression levels for G/G are more consistent compared to A/A.
There are some outliers in both groups, but especially in the A/A group.
These patterns suggest the A/A genotype may be linked to increased ORMDL3 expression compared to G/G.

