---
title: "Class6"
author: "Erin Li"
format: pdf
---

```{r}
student1<-c(100,100,100,100,100,100,100,90) 
student2<-c(100,NA,90,90,90,90,97,80) 
student3<-c(90,NA,NA,NA,NA,NA,NA,NA)
```

```{r}
mean(student1)
mean(student2)
mean(student3)
```

```{r}
mean(student1[-which.min(student1)])
mean(student2[-which.min(student2)])
mean(student3[-which.min(student3)])
#which.min(student1) gives position 8

```

```{r}
mean(student2,na.rm = TRUE)
mean(student3, na.rm = TRUE)
```


I want to stop working with 'student1', 'student2' etc. and typing it out every time so let instead work with an input called 'x'
```{r}
#Mask NA values to zero
x <- student2
x [is.na(x)]<-0
#Drop lowest score and get the mean
result <- mean(x[-which.min(x)])
result
```
Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
grade <- function(x) {
  #Mask NA values to zero
  x [is.na(x)]<-0
  #Drop lowest score and get the mean
  mean(x[-which.min(x)])
}
```

Use this function:
```{r}
grade(student1)
grade(student2)
grade(student3)
```

```{r}
gradebook<-read.csv('https://tinyurl.com/gradeinput', row.names=1)
gradebook
```
Q2.Usingyourgrade()functionandthesuppliedgradebook,Whoisthetopscoringstudent overallinthegradebook?[3pts]
Top scoring student is 18, scored 94.5.
```{r}
ans = apply(X= gradebook, MARGIN=1, FUN=grade)
ans
#which.max(list)
#ans[which.max(list)]
#Top scoring student is 18, scored 94.5.
```

Q3.From your analysis of the gradebook,which homework was toughest on students(i.e.obtained thelowestscoresoverall?[2pts]
Question 2 is the toughest.
```{r}
mask <- gradebook
mask[is.na(mask)]<-0
hw.ave <- apply(mask, 2, mean)
hw.ave
# Question 2 is the toughest
```
Q4.OptionalExtension:Fromyouranalysisofthegradebook,whichhomeworkwasmost predictiveofoverallscore(i.e.highestcorrelationwithaveragegradescore)?[1pt]
hw 5 is the most representative
```{r}
mask <- gradebook
mask[is.na(mask)]<-0
hw.ave <- apply(mask, 2, mean)
correlate <- apply(mask, 2, cor, y = ans)
correlate
```

```{r}
which.max(correlate)
#hw 5 is the most representative
```

